---
title: "ExpSigfinder matrix correction"
author: "Ellie Dunstone"
date: "2023-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230127/extractions/wp4_2530_20230324_expsigfinder_SBS96/")
```

```{r packages, include=FALSE}
##load packages
library(tidyverse)
```

## Introduction 

This is a script for taking a mutation matrix generated from NanoSeq sequencing of samples exposed to mutagens in vitro, subsetting it for samples that show significantly increased mutagenesis compared to control (as identified by expsigfinder_mut_detection.Rmd), and then correcting it to account for variation in duplex coverage.

This script is part of ExpSigfinder-nano (https://github.com/EllieDunstone/ExpSigfinder-nano), based on functions written by Zoe Zou (https://github.com/xqzou/ExpSigfinder). It should be run before the scripts for extracting signatures, in order to correct the input matrix.


## Read in uncorrected mutation matrix

Read in the uncorrected mutation matrix. This needs to be formatted with sample names in the "sample" column, and mutation classes as subsequent columns, with the counts of each mutation class in each sample as the matrix values.

```{r read matrix}
mutation_matrix <- read.table("/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230127/data/trint_subs_matrix_corrected.tsv", sep = "\t", header = T, as.is = T) 
```

If required, trim the sample names (sometimes these get output with the parent dir prefixed).

```{r trim sample names}
#Trim sample IDs 
colnames(mutation_matrix)[-1] <- stringr::str_sub(colnames(mutation_matrix), 30, -1)[-1] #May need to alter the start position depending on input
```


## Read in the corrected mutation burden data

Read in the corrected mutation burden data for making the corrections.

```{r read corrected burdens}
burdens <- read.table("/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230127/data/results.mut_burden_all.tsv", sep = "\t", header = T, as.is = T) 
```


## Read in mutagenesis tables

Read in the mutagenesis tables output by expsigfinder_mut_detection.Rmd

```{r read mutagenesis tables}
#list mut tables in working dir
mut_tables_list <- list.files(pattern = "mutagenesis_table")

# initialize an empty dataframe 
mutagenesis_table <- data.frame()

# loop over the file names
for (file in mut_tables_list) {
  # read in the current file and assign it to a dataframe
  df <- read.table(paste0(getwd(), "/", file), header = TRUE, sep = ",")
  
  # concatenate the dataframe with the existing data
  mutagenesis_table <- rbind(mutagenesis_table, df)
}

```

We can now subset this table to contain only samples with evidence of significant increase in mutagenesis, along with control samples:

```{r subset mutagenesis table}
sign_mutagenesis_table <- filter(mutagenesis_table, sign_mutagenesis == "Y" | group == "control")
```


## Subset uncorrected matrix

We can now subset the input matrix to contain only samples with evidence of significant increase in mutagenesis.

```{r subset mutation matrix}
sign_mutation_matrix <-  mutation_matrix[names(mutation_matrix) %in% sign_mutagenesis_table$sample]

#add back first column
sign_mutation_matrix <- cbind(mutation_matrix[,1], sign_mutation_matrix)
colnames(sign_mutation_matrix)[1] <- "MutationType"
```


## Assess total mutation call numbers

First, we want to calculate the total number of mutation calls per sample

```{r sum mut calls}
#transpose matrix to make calculation table
mut_matrix_calcs_table <- t(sign_mutation_matrix[, -1])
colnames(mut_matrix_calcs_table) <- sign_mutation_matrix[, 1]

#calculate sums
df_sum <- mut_matrix_calcs_table %>% as.matrix() %>% rowSums()
mut_matrix_calcs_table <- mutate(as.data.frame(mut_matrix_calcs_table), total_muts = df_sum)
```

We can now plot these to assess whether there are sufficient mutation calls in the samples to assume that noise in the mutation spectra is reasonably negligible.

```{r plot total mut calls}

total_mut_calls_hist <- ggplot(mut_matrix_calcs_table, aes(x = total_muts)) + 
  geom_boxplot(fill = "hotpink") + 
  labs(title = paste0("Range: ", min(mut_matrix_calcs_table$total_muts), "-", max(mut_matrix_calcs_table$total_muts)))

total_mut_calls_hist
```

We can set a threshold below which to drop samples - these samples may have evidence for excess mutagenesis, but insufficient raw mutation calls to assess the signature. Note: should this be changed to prevent dropping of any controls?

```{r drop low mut calls}
#identify samples below threshold and drop from calculation table, output dropped samples
mut_matrix_calcs_table_keep <- filter(mut_matrix_calcs_table, total_muts >= 200)
mut_matrix_calcs_table_dropped <- filter(mut_matrix_calcs_table, total_muts < 200)
write.csv(mut_matrix_calcs_table_dropped, "mutation_matrix_low_raw_calls.csv")

#drop from the mutation matrix
sign_mutation_matrix <-  sign_mutation_matrix[names(sign_mutation_matrix) %in% rownames(mut_matrix_calcs_table_keep)]

#add back first column
sign_mutation_matrix <- cbind(mutation_matrix[,1], sign_mutation_matrix)
colnames(sign_mutation_matrix)[1] <- "MutationType"
```


## Correct matrix

We can now correct the matrix to the estimated number of mutations per diploid human genome.

```{r correct matrix}
#Add total duplex coverage to calculation table and calculate correction factor
mut_matrix_calcs_table_keep <- mut_matrix_calcs_table_keep %>% mutate(sample = rownames(mut_matrix_calcs_table_keep)) %>% left_join(select(burdens, sample, total)) %>% mutate(cf = 6000000000/total)

#check order
mut_matrix_calcs_table_keep["sample"] == colnames(sign_mutation_matrix)[-1]

#Multiply matrix by factors
corrected_counts <- sign_mutation_matrix[,-1]*t(mut_matrix_calcs_table_keep["cf"])
sign_mutation_matrix_coverage_corrected <- cbind(sign_mutation_matrix[,1], round(corrected_counts))
colnames(sign_mutation_matrix_coverage_corrected)[1] <- "MutationType"
```

And output it for use in downstream analysis.

```{r output corrected matrix}
write.csv(sign_mutation_matrix_coverage_corrected, "sign_mutation_matrix_coverage_corrected.csv", row.names = FALSE)
```

