---
title: "ExpSigfinder wrapper SBS"
author: "Ellie Dunstone"
date: "15/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230406/extractions/wp4_2530_expsigfinder_SBS96_test_boundary//")
```

```{r packages, include=FALSE}
##load packages
library(tidyverse)
#library(ExpSigfinder)
library(ExpSigfinder-nano)
library(Rmisc)
library(ggrepel)
library(data.table)
```

## Introduction 

This is a script to implement ExpSigfinder-nano (https://github.com/EllieDunstone/ExpSigfinder-nano), based on functions written by Zoe Zou (https://github.com/xqzou/ExpSigfinder). These scripts are for analysing experimental mutational signature data from in vitro models sequenced using the NanoSeq protocol (Abascal et al., 2021). 

## Install package

You only need to run this the first time to install the github package. If you need to do this, then re run the package loading chunk above afterwards.

```{r install expsigfinder}
##install package, only needed when running for the first time
library(devtools)
install_github("xqzou/ExpSigfinder")
#install_github("EllieDunstone/ExpSigfinder-nano")
```

## Read in mutation matrix

Read in the mutation matrix (input your own path to file). This needs to be formatted with sample names in the "sample" column, and mutation classes as subsequent columns, with the counts of each mutation class in each sample as the matrix values. This matrix should contain samples with evidence of a significant increase in mutagenesis compared to the control samples (see expsigfinder_mut_detection.Rmd) plus the control samples, and be corrected for duplex coverage (see expsigfinder_matrix_correction.Rmd).

```{r read matrix}
mutation_matrix <- read.table("sign_mutation_matrix_coverage_corrected.csv", sep = ",", header = T, as.is = T) 
```

If required, trim the sample names (sometimes these get output with the parent dir prefixed).

```{r trim sample names}
#Trim sample IDs 
colnames(mutation_matrix)[-1] <- stringr::str_sub(colnames(mutation_matrix), 30, -1)[-1] #May need to alter the start position depending on input
```

## Read in metadata

Read in table of metadata containing the treatment name, and a column 'group' indicating whether a sample is a control. Other metadata columns may be included if needed for later analysis.

```{r read metadata}
metadata <- read.table("/Users/ed4/Documents/phd/chemo_project/Metadata/WP4/wp4_sample_subsetting.csv", sep = ",", header = T, as.is = T)
```


## Subset per organoid line

Background needs to be calculated per organoid line, this chunk creates a mutation matrix for each line.

In order to deal with the difference between the KO and WT iPSC lines, we will have to append to the patient name to differentiate them:

```{r differentiate WT and KO iPSC lines}

#fix metadata table
for (i in 1:nrow(metadata)) {
  if (any(grepl("cBOB", metadata[i,]))) {
    if (any(grepl("KO", metadata[i,]))) {
      metadata$patient[i] <- "PD44571_KO"
      metadata$sample[i] <- paste0(substr(metadata$sample[i], 1, 7), "_KO", substr(metadata$sample[i], 8, nchar(metadata$sample[i])))
    } else {
      metadata$patient[i] <- "PD44571_WT"
      metadata$sample[i] <- paste0(substr(metadata$sample[i], 1, 7), "_WT", substr(metadata$sample[i], 8, nchar(metadata$sample[i])))
    }
  }
}

#fix mutation matrix
for (i in 2:ncol(mutation_matrix)) {
  if (grepl("PD44571[i-n]", colnames(mutation_matrix)[i])) {
    colnames(mutation_matrix)[i] <- paste0(substr(colnames(mutation_matrix)[i], 1, 7), "_KO", substr(colnames(mutation_matrix)[i], 8, nchar(colnames(mutation_matrix)[i]))) 
  }
  if (grepl("PD44571[o-v]", colnames(mutation_matrix)[i])) {
    colnames(mutation_matrix)[i] <- paste0(substr(colnames(mutation_matrix)[i], 1, 7), "_WT", substr(colnames(mutation_matrix)[i], 8, nchar(colnames(mutation_matrix)[i]))) 
  }
}
```


```{r subset matrix per donor}
org_lines <- unique(metadata$patient) #grab all organoid line IDs


for (org in org_lines) {
  assign(paste0(org, "_mutation_matrix"), cbind(MutationType = mutation_matrix[,1], dplyr::select(mutation_matrix, matches(org))))
}

```


## Add metadata

Create a dataframe using the relevant metadata columns and the mutation matrix. This is currently at a per line level, select your organoid line first here:

```{r select org line}
org <- "PD57287"
```

Create a dataframe using the relevant metadata columns and the mutation matrix. This is currently at a per line level, could be automated to do all at once?

```{r add metadata}

t_mutation_matrix <- t(get(paste0(org, "_mutation_matrix")))
colnames(t_mutation_matrix) <- t_mutation_matrix[1,]
t_mutation_matrix <- t_mutation_matrix[-1,]
t_mutation_matrix <- as.data.frame(cbind(rownames(t_mutation_matrix), t_mutation_matrix))
colnames(t_mutation_matrix)[1] <- "sample"

mutation_matrix_metadata <- dplyr::left_join(t_mutation_matrix, dplyr::select(metadata, sample, patient, tissue_sample, treatment, sample_name, concentration, tissue, group))
```

## Calculate background

Calculate the means of control samples for each mutation type, and the overall mean burden. 

First, we identify the control samples:

```{r select controls}
controls <- filter(mutation_matrix_metadata, group == "Control")
control_names <- controls$sample
```

If you need to exclude any problematic controls at this stage, this chunk can be used:

```{r adjust controls}
controls <- filter(mutation_matrix_metadata, group=="Control" & sample!="PD52491x_ds0001" & sample!="PD52491u_ds0001")
control_names <- control_names[-c(6,7)]
```

We now calculate the background burden and spectrum using these samples.

```{r calculate background}
#calculate mean counts of each mutation type across all control samples
PD57287_mutation_matrix <- PD57287_mutation_matrix %>%
  mutate(bg_profile = rowMeans(PD57287_mutation_matrix[,control_names]))

#calculate mean total mutation burden across all control samples
bg_mean <- sum(PD57287_mutation_matrix$bg_profile)
print(bg_mean)
```


## Make profiles for treatment groups and plot


```{r calculate and plot treatment group profiles - individual treatment group}

#Note: Currently will not output plots to correct location - needs fixing

#select appropriate samples by treatment group
treatment_group <- filter(mutation_matrix_metadata, treatment == "thiotepa")
group_names <- treatment_group$sample

# Make profile of treatment group samples with background removed (input sample names here)
Wrap_KOSig(PD41850_mutation_matrix, "bg_profile", group_names, 100, bg_mean, 2, "thiotepa")
sig <- read.table("thiotepa.txt", sep = "\t", header = T, as.is = T)
plotCountbasis(sig, 1, 6, 9, paste0("thiotepa", ".pdf"))
plotPercentagebasis(sig, 1, 6, 9, paste0("thiotepa", "_percentage.pdf"))
```

Automated for all treatment groups:

```{r calculate and plot treatment group profiles for all treatment groups}

#Generate list of non-control treatment group names
treatment_groups <- unique(filter(mutation_matrix_metadata, group != "Control")$treatment)

#Run calculation and plotting for each treatment group
for (i in treatment_groups) {
  print(paste0("Analysing treatment group: ", i))
  
  #select appropriate samples by treatment group
  treatment_group <- filter(mutation_matrix_metadata, treatment==i)
  group_names <- treatment_group$sample
  print(paste0(group_names))
  
  # Make profile of treatment group samples with background removed - skip and print error if one group fails
  tryCatch({
    Wrap_KOSig(PD41849_mutation_matrix, "bg_profile", group_names, 100, bg_mean, 2, i)
    sig <- read.table(paste0(i,".txt"), sep = "\t", header = T, as.is = T)
    plotCountbasis(sig, 1, 6, 9, paste0(i, ".pdf"))
    plotPercentagebasis(sig, 1, 6, 9, paste0(i, "_percentage.pdf"))}, error=function(e){cat(paste0("ERROR for treatment group ", i, ": "),conditionMessage(e), "\n")})
}

```

Or for something other than treatment group...

```{r calculate and plot group profiles for custom groups}

#Generate list of non-control group names
custom_groups <- unique(filter(mutation_matrix_metadata, group != "control")$tissue)

#Run calculation and plotting for each treatment group
for (i in custom_groups) {
  print(paste0("Analysing group: ", i))
  
  #select appropriate samples by treatment group
  custom_group <- filter(mutation_matrix_metadata, tissue==i)
  group_names <- custom_group$sample
  print(paste0(group_names))
  
  # Make profile of treatment group samples with background removed - skip and print error if one group fails
  tryCatch({
    Wrap_KOSig(PD53370_mutation_matrix, "bg_profile", group_names, 100, bg_mean, 2, i)
    sig <- read.table(paste0(i,".txt"), sep = "\t", header = T, as.is = T)
    plotCountbasis(sig, 1, 6, 9, paste0(i, ".pdf"))
    plotPercentagebasis(sig, 1, 6, 9, paste0(i, "_percentage.pdf"))}, error=function(e){cat(paste0("ERROR for group ", i, ": "),conditionMessage(e), "\n")})
}

```

Or for individual samples:

```{r calculate and plot profiles for all samples individually}

#Generate list of non-control sample names
non_control_samples <- filter(mutation_matrix_metadata, group != "Control")$sample

#Run calculation and plotting for each sample
for (i in non_control_samples) {
  print(paste0("Analysing sample: ", i))

  # Make profile of sample with background removed - skip and print error if one sample fails
  tryCatch({
    Wrap_KOSig(PD57287_mutation_matrix, "bg_profile", i, 100, bg_mean, 2, i)
    sig <- read.table(paste0(i,".txt"), sep = "\t", header = T, as.is = T)
    plotCountbasis(sig, 1, 6, 9, paste0(i, ".pdf"))
    plotPercentagebasis(sig, 1, 6, 9, paste0(i, "_percentage.pdf"))}, error=function(e){cat(paste0("ERROR for sample ", i, ": "),conditionMessage(e), "\n")})
}

```




## Calculate mean burden per genome and confidence intervals for control

```bg_mean``` is the mean number of raw mutation calls per genome, but we need to account for variation in effective coverage (total base calls). 

First we select the controls for this patient:

```{r select controls}
controls_stats <- filter(metadata, sample %in% control_names & patient=="PD57287")
mean(controls_stats$muts) # not entirely clear on why this is slightly higher than bg_mean?
```

Then calculate the genome burden (burden per base x diploid genome size in bp), and plot and test its distribution:

```{r calculate genome burden}
#add genome burden per sample
controls_stats <- mutate(controls_stats, genome_burden = burden*6000000000)
metadata <- mutate(metadata, genome_burden = burden*6000000000)

#plot distribution of genome burdens - at some point do this across all patients
genome_burden_plot <- ggplot(controls_stats, aes(genome_burden)) +
  geom_boxplot() +
  geom_dotplot(dotsize = 5, binwidth = 1) +
  geom_text_repel(aes(y=0, label=treatment))

genome_burden_plot # some 'outliers' - how different are the spectra? does this matter?


#Summary stats for controls
control_summary_stats <- metadata %>% filter(sample %in% control_names & patient=="PD57287") %>% dplyr::group_by(group) %>%
   dplyr::summarise(mean=mean(genome_burden), max=max(genome_burden), min=min(genome_burden), median=median(genome_burden), sd=sd(genome_burden), count=n())

CI_table_controls<-as.data.table(filter(metadata, sample %in% control_names & patient=="PD57287"))[,as.list(CI(genome_burden)), by=group]
control_summary_stats <-left_join(control_summary_stats, select(CI_table_controls,-c(mean)),by="group")

#Summary stats per treatment group
treatment_group_summary_stats <- metadata %>% filter(group!="control" & patient=="PD57287") %>%
  dplyr::group_by(treatment) %>% dplyr::summarise(mean=mean(genome_burden), max=max(genome_burden), min=min(genome_burden), median=median(genome_burden), sd=sd(genome_burden), count=n())

CI_table_treatments <- as.data.table(filter(metadata, group!="control" & patient=="PD57287"))[,as.list(CI(genome_burden)), by=treatment]
treatment_group_summary_stats <-left_join(treatment_group_summary_stats, select(CI_table_treatments,-c(mean)),by="treatment")

#Summary stats per custom group
custom_group_summary_stats <- metadata %>% filter(group!="control" & patient=="PD44571") %>%
  dplyr::group_by(tissue) %>% dplyr::summarise(mean=mean(genome_burden), max=max(genome_burden), min=min(genome_burden), median=median(genome_burden), sd=sd(genome_burden), count=n())

CI_table_custom <- as.data.table(filter(metadata, group!="control" & patient=="PD44571"))[,as.list(CI(genome_burden)), by=tissue]
custom_group_summary_stats <-left_join(custom_group_summary_stats, select(CI_table_custom,-c(mean)),by="tissue")

```

## Combine signature outputs

We can now combine the output signatures into a table for downstream analysis.

```{r generate sig table, message=FALSE}
parent_dir <- "/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230406/extractions/wp4_2530_20230411_expsigfinder_SBS96/"
org_line_dirs <- list.files(parent_dir, pattern = "PD")

#init dataframe
all_sigs <- as.data.frame(matrix(nrow = 96, ncol = 1))

#get list of sig files
sig_files_list <- list.files(pattern = ".txt")


for (file in sig_files_list) {
    sig_file <- read_table(file)
    name <- str_split_i(file, "\\.", 1)
    print(name)
    sig_file <- mutate(sig_file, norm_exposure = KO_exposure / sum(KO_exposure)) #normalise signature to sum to 1
    all_sigs <- cbind(all_sigs, sig_file$norm_exposure) 
    colnames(all_sigs)[length(colnames(all_sigs))] <- name
  }

all_sigs[1] <- sig_file$MutationType
colnames(all_sigs)[1] <- "MutationType"


write_csv(all_sigs, "expsigfinder_bg_removed_all_sigs_SBS96.csv")
```

If you need to debug, output a list of all samples analysed and whether or not they are succeeding:

```{r output debug table}
debug_table <- as.data.frame(cbind(sample = colnames(mutation_matrix)[-1], working = NA))
debug_table <- mutate(debug_table, working = case_when(sample %in% colnames(all_sigs) ~ "Y", TRUE ~ "N"))
debug_table <- left_join(debug_table, metadata)

write_csv(debug_table, "expsigfinder_debug_table.csv")
```


## Next steps

Make sure I understand the mathematical principles behind these calculations and exactly which numbers are being output in the summary table for each treatment group.

Work out why some of the treatment groups are currently failing.


Work out how to decide which controls to use - are these sufficiently similar in spectrum and burden within each org line?


Downstream analysis to determining whether the resulting signatures are significantly different between:
  - concentrations, +-S9, etc.
  - Treatment groups: e.g. cisplatin vs. carboplatin
  - Organoid lines: e.g. bendamustine in PD41849 vs. bendamustine in PD41850
  - Tissues of origin: e.g. is there evidence for a significant effect of tissue of origin, after accounting for the effect of donor of origin


Error bars or other estimates of signature "confidence"?

Once we have the final set of signatures:
  - Comparison to COSMIC signatures and previous iPSC signatures - will probably attempt decomposition using sigfit.
  - Compare to the results obtained using de novo extraction methods.
  
Implement for indels and DBS (although likely to be limited to very strong signatures due to lower mutation numbers).





