---
title: "expsigfinder_mut_detection.Rmd"
author: "Ellie Dunstone"
date: "2023-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230127/extractions/wp4_2530_20230324_expsigfinder_SBS96/")
```

```{r packages, include=FALSE}
##load packages
library(tidyverse)
library(NatParksPalettes) #Not essential, colour palette for plotting - other palette can be used
```

## Introduction 

This is a script for identifying in vitro samples with a significantly higher mutation burden than corresponding control samples. It is designed for use with samples sequenced using NanoSeq.

This script is part of ExpSigfinder (https://github.com/EllieDunstone/ExpSigfinder), based on functions written by Zoe Zou (2022). It should be run before the scripts for extracting signatures, in order to define the samples where there is evidence of significant mutation increase above the background. 

## Read in mutation matrix

Read in the mutation matrix. This needs to be formatted with sample names in the "sample" column, and mutation classes as subsequent columns, with the counts of each mutation class in each sample as the matrix values.

```{r read matrix}
mutation_matrix <- read.table("/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230127/data/trint_subs_matrix_corrected.tsv", sep = "\t", header = T, as.is = T) 
```

If required, trim the sample names (sometimes these get output with the parent dir prefixed).

```{r trim sample names}
#Trim sample IDs 
colnames(mutation_matrix)[-1] <- stringr::str_sub(colnames(mutation_matrix), 30, -1)[-1] #May need to alter the start position depending on input
```


## Read in the corrected mutation burden data

Read in the corrected mutation burden data for identification of samples showing significant increase compared to the background.

```{r read corrected burdens}
burdens <- read.table("/Users/ed4/Documents/phd/chemo_project/BotSeq/WP4/KCL_organoid_work/wp4_2530_20230127/data/results.mut_burden_all.tsv", sep = "\t", header = T, as.is = T) 
```


## Read in metadata

Read in table of metadata containing the treatment name, and a column 'group' indicating whether a sample is a control. Other metadata columns may be included if needed for later analysis.

```{r read metadata}
metadata <- read.table("/Users/ed4/Documents/phd/chemo_project/Metadata/WP4/wp4_sample_subsetting.csv", sep = ",", header = T, as.is = T)
```


## Subset per organoid line

Background needs to be calculated per organoid line, this chunk creates a mutation matrix for each line.

```{r subset matrix per donor}
org_lines <- unique(metadata$patient) #grab all organoid line IDs

for (org in org_lines) {
  assign(paste0(org, "_mutation_matrix"), cbind(MutationType = mutation_matrix[,1], dplyr::select(mutation_matrix, matches(org))))
}
```

## Add metadata

Create a dataframe using the relevant metadata columns and the mutation matrix. This is currently at a per line level, could be automated to do all at once? For now, select your organoid line first here:

```{r select org line}
org <- "PD41850"
```


```{r add metadata}

t_mutation_matrix <- t(get(paste0(org, "_mutation_matrix")))
colnames(t_mutation_matrix) <- t_mutation_matrix[1,]
t_mutation_matrix <- t_mutation_matrix[-1,]
t_mutation_matrix <- as.data.frame(cbind(rownames(t_mutation_matrix), t_mutation_matrix))
colnames(t_mutation_matrix)[1] <- "sample"

mutation_matrix_metadata <- dplyr::left_join(t_mutation_matrix, dplyr::select(metadata, sample, patient, tissue_sample, treatment, sample_name, concentration, tissue, group))
```

```{r append burdens}
mutation_matrix_metadata_burdens <- left_join(mutation_matrix_metadata, burdens)
```

## Calculate distribution of burdens for controls

Calculate the means of control samples for each mutation type, and the overall mean burden.

First, we identify the control samples:

```{r select controls}
controls <- filter(mutation_matrix_metadata_burdens, group == "control")
control_names <- controls$sample
```

Plot the distribution of the control burdens:

```{r plot control burdens}
control_burdens_plot <- ggplot(controls, aes(reorder(sample_name, burden), burden, fill = treatment)) +
  geom_col() +
  geom_linerange(aes(x = sample_name, ymin = burden_lci, ymax = burden_uci)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Sample") +
  ylab("SBS burden per bp") + 
  facet_grid(~treatment, scales = "free") +
  scale_fill_manual(values = natparks.pals("Yellowstone", n = length(unique(controls$treatment))))
  
control_burdens_plot
```

Do we want to add any formal statistical test here? We have effectively the sample burdens, and upper and lower confidence intervals. For some pairs, the confidence intervals don't quite overlap, but there certainly doesn't look like there is any systematic difference so this is broadly unconcerning? (For PD41849)

If there is no evidence of meaningful differences between the control samples, they can be treated as replicates of the control/background burden and spectrum. 

We now calculate the mean background mutation count and spectrum using these samples.

```{r calculate background counts and spectrum}
#calculate mean counts of each mutation type across all control samples
PD41849_mutation_matrix <- PD41849_mutation_matrix %>%
  mutate(bg_profile = rowMeans(PD41849_mutation_matrix[,control_names]))

#calculate mean total mutation count across all control samples
bg_mean_count <- sum(PD41849_mutation_matrix$bg_profile)
print(bg_mean_count)
```

We now want to calculate the mean and confidence interval of corrected burdens for the controls.

```{r calculate background burden mean and CI}
#Summary stats for bg burdens
bg_summary_stats <- metadata %>% filter(sample %in% control_names & patient=="PD41849") %>% dplyr::group_by(group) %>%
   dplyr::summarise(mean = mean(burden), max = max(burden), min = min(burden), median = median(burden), sd = sd(burden), count = n())

#Add 95% CIs to bg burden summary stats
bg_CI_table <- as.data.table(filter(metadata, sample %in% control_names & patient == "PD41849"))[,as.list(CI(burden)), by = group]
bg_summary_stats <-left_join(bg_summary_stats, select(bg_CI_table, -c(mean)), by = "group")
```


#Identify samples with significantly increased burdens

We can now use these values to identify which individual samples show evidence of significant increase in burden.

First, we can plot the same plot as above with the control distributions, but adding the samples for each treatment to visually inspect each group for evidence of increased burden.

```{r plot control burdens alongside treatment}

#generate list of all non-control treatments
treatment_list <- unique(filter(mutation_matrix_metadata_burdens, group != "control")$treatment)

for (i in treatment_list) {
  control_v_treatment_burdens_plot <- ggplot(filter(mutation_matrix_metadata_burdens, patient == "PD41849" & (group == "control" | treatment == i)), aes(reorder(sample_name, -burden), burden, fill = group)) + geom_col() +
  geom_linerange(aes(x = sample_name, ymin = burden_lci, ymax = burden_uci)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), strip.text.x = element_text(angle = 90)) +
  xlab("Sample") +
  ylab("SBS burden per bp") + 
  labs(title = paste0("Control 95% CI: ", signif(bg_summary_stats$lower, 3), " - ", signif(bg_summary_stats$upper, 3))) +
  #annotate(geom = "text", x = 0, y = max(filter(mutation_matrix_metadata_burdens, patient == "PD41849" & (group == "control"| treatment == i))$burden_uci), label = paste0("Control 95% CI: ", bg_summary_stats$lower, " - ", bg_summary_stats$upper), nudge_y = 0.1, data = filter(mutation_matrix_metadata_burdens, treatment == i), size = 3, hjust = 0) +
  facet_grid(~treatment, scales = "free") +
  scale_fill_manual(values = natparks.pals("Volcanoes", n = 2))
  
ggsave(paste0("PD41849_controls_v_", i, ".png"))
}


```

We also want to output these results to a table.

We can add whether there is evidence of significant excess mutagenesis in each sample.

The sign_mutagenesis column will be 'Y' if the burden is significantly higher than the background, 'N' if it lies within the 95% CI of the background, and 'low' if significantly lower. 'Low' results are unusual, as the treatment should not significantly decrease the burden! Some borderline 'low' results may represent some experimental variation, but large apparent decreases should be investigated further for sources of error.

```{r generate table of mutagenesis results}
#select data from current table
mutagenesis_table <- select(mutation_matrix_metadata_burdens, sample, patient, tissue_sample, treatment, sample_name, concentration, tissue, group, muts, total, burden, burden_lci, burden_uci)

#add bg summary stats
mutagenesis_table <- mutagenesis_table %>% mutate(bg_burden_mean = bg_summary_stats$mean) %>% mutate(bg_burden_lci = bg_summary_stats$lower) %>% mutate(bg_burden_uci = bg_summary_stats$upper)

#add whether sample CI overlaps bg CI - 'Y' if higher, 'N' if within bg CI, low if below bg CI - if this occurs this is odd and should be investigated
mutagenesis_table <- mutagenesis_table %>% mutate(sign_mutagenesis = case_when(burden_lci >= bg_burden_uci ~ "Y", (burden_lci < bg_burden_uci) & (burden_uci >= bg_burden_lci) ~ "N", TRUE ~ "low" ))

```


# Identifying samples of interest

The simplest way to identify samples of interest is by simply looking at all samples that show a significant increase:

```{r subset samples - sign increase}
samples_of_interest <- filter(mutagenesis_table, sign_mutagenesis == "Y")$sample
samples_of_interest
```

However, you may also want to add or remove samples from this set manually:

```{r subset samples - manual}
samples_of_interest <- cbind(filter(mutagenesis_table, sign_mutagenesis == "Y")$sample, c("your_sample", "another_one"))
samples_of_interest
```


We can now generate a data table for your samples of interest:

```{r data table samples of interest}
data_samples_of_interest <- filter(mutation_matrix_metadata_burdens, sample %in% samples_of_interest)
```

And also inspect the other samples that did not show significant mutagenesis:

```{r data table samples of interest}
data_dropped_samples <- filter(mutation_matrix_metadata_burdens, !(sample %in% samples_of_interest))
```


# Output tables

We can now output these tables for use downstream:

```{r output data tables}
write_csv(data_samples_of_interest, paste0(org, "_data_samples_of_interest.csv"))
write_csv(data_dropped_samples, paste0(org, "_data_dropped_samples.csv"))
```

